---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# fidelioDiagnostics

`fidelioDiagnostics` is an R package that:

- extracts **FIDELIO** model results from GDX files,
- converts them to wide tables and computes **derived indicators** (nation, industry, trade),
- saves outputs per symbol (Parquet/Feather/…) plus a **manifest** and optional **bundles**,
- and ships two Shiny apps: an internal **diagnostic** app and a shareable **results** app.

---

## Installation

```{r, eval=FALSE}
# install the dev version from GitHub (edit owner/org as needed)
# install.packages("remotes")
remotes::install_github("your-org/fidelioDiagnostics")
```

For local development:

```{r, eval=FALSE}
# from package root
# install.packages("devtools")
devtools::load_all()      # develop interactively (no install)
devtools::document()      # update docs/NAMESPACE from roxygen comments
devtools::install()       # build & install into your R library
```

---

## Configure

Edit `config/project.yml`. Minimal example:

```yaml
paths:
  gdx_dir: "data-raw/gdx/eta_cpi_025_beta_infl_1"
  outputs: "outputs"

scenarios: ["baseline","ff55"]

groups:
  EU28: ["AUT","BEL","BGR","HRV","CYP","CZE","DNK","EST","FIN","FRA","DEU","GRC",
         "HUN","IRL","ITA","LVA","LTU","LUX","MLT","NLD","POL","PRT","ROU",
         "SVK","SVN","ESP","SWE","GBR"]

save:
  formats: ["parquet","feather"]   # per-symbol files
  bundles:
    diagnostic_app:
      include: ["GDPr_t","HDY_VAL_t","HDYr_t","I_TOT_PP_t","DS_t","FS_t","GSUR_VAL_t",
                "GINV_VAL_t","TBr_t","TB_GDP_t","HSAVR_t","U_t","KLratio_country_t",
                "ir_t","P_HH_CPI_t","I_PP_t","K_t","L_t","GHG_t","KLratio_t",
                "P_Q_t","P_KL_t","I_PP_SECT6_t","OUT_COMP6_SHARE_REAL_t","BITRADE_REG_t"]

    results_app:
      include: ["GDPr_t","TBr_t","TB_GDP_t","I_PP_SECT6_t","OUT_COMP6_SHARE_REAL_t","BITRADE_REG_t"]
      filters:
        BITRADE_REG_t:
          keep:
            c: ["TOT"]             # keep only commodity total
      csv_combine: true
      csv_basename: "results_bundle"
      csv_shape: "wide"
```

---

## Quick start

```{r, eval=FALSE}
library(fidelioDiagnostics)

# 1) Load config and print paths
cfg <- load_config()
print_runtime_info(cfg)

# 2) Run the pipeline: extract → wide → aggregates → derived → save
res <- run_pipeline()

# 3) Outputs (by default):
# - outputs/derived/<symbol>.parquet (and/or .feather/.csv/.rds per config)
# - outputs/derived/manifest.rds (index for lazy loading in apps)
# - outputs/derived/bundle_<name>.rds for configured bundles
# - outputs/derived/results_bundle.csv if enabled for the results bundle
```

---

## Launch the apps

```{r, eval=FALSE}
# Internal diagnostics (full set)
launch_app("diagnostic")

# Slim results app (subset; uses manifest or bundle_results_app.rds)
launch_app("results")
```

Both apps **lazy-load** tables using `outputs/derived/manifest.rds`. If present, they can also load targeted `bundle_*.rds`.

---

## What’s produced

- **Nation (n,t)**: `GDPr_t`, `HDY_VAL_t`, `HDYr_t`, `I_TOT_PP_t`, `DS_t`, `FS_t`,
  `GSUR_VAL_t`, `GINV_VAL_t`, `TBr_t`, `TB_GDP_t`, `HSAVR_t`, `U_t`,
  `KLratio_country_t`, `ir_t`, `P_HH_CPI_t`.
- **Industry (n,i,t)**: `I_PP_t`, `K_t`, `L_t`, `GHG_t`, `KLratio_t`, `P_Q_t`,
  `P_KL_t`, `I_PP_SECT6_t`, `OUT_COMP6_SHARE_REAL_t`.
- **Bilateral trade (n,n1,c,t)**: `BITRADE_REG_t` (with totals rows appended).

Tables are wide by scenario (`baseline`, `ff55`) and include `delta` and `pct` where relevant.

---

## Developer quick guide (devtools)

### What each command does

- `devtools::document()`  
  Generate **NAMESPACE** and **man/*.Rd** from roxygen comments (`#' @export`, `@param`, …).

- `devtools::load_all()`  
  Load the package **from source** into the current session (no install). Fast feedback loop.

- `devtools::install()`  
  Build & install into your user library (use in fresh R sessions or scripts).

- `devtools::check()`  
  Run CRAN-like checks (R CMD check, docs, examples, namespace, DESCRIPTION, etc.).

- `devtools::build()`  
  Create a source tarball (`.tar.gz`) you can share/install elsewhere.

- `devtools::test()`  
  Run **testthat** tests under `tests/`.

- `devtools::build_vignettes()` / `devtools::clean_vignettes()`  
  Build/clean vignettes if you ship them.

- `devtools::build_readme()`  
  Knit `README.Rmd` → `README.md`.

> **Tip:** If you want roxygen to manage `NAMESPACE`, delete any manually-created `NAMESPACE` and run `devtools::document()`.

### Common recipes

#### 1) Fast inner dev loop

    # edit code in R/*.R
    devtools::load_all()

    # if you changed roxygen tags/exports/docs:
    devtools::document()
    devtools::load_all()

    # try your functions/pipeline/apps (same session)
    fidelioDiagnostics::run_pipeline()
    fidelioDiagnostics::launch_app("diagnostic")  # or "results"

#### 2) Install & use in a fresh session

    devtools::document()
    devtools::install(upgrade = "never", dependencies = FALSE, build_vignettes = FALSE)

    # restart the R session, then:
    .rs.restartR(); library(fidelioDiagnostics)
    fidelioDiagnostics::launch_app("results")

#### 3) Pre-commit / pre-share sanity check

    devtools::document()
    devtools::check()

#### 4) Build a distributable (source tarball)

    devtools::document()
    devtools::check()
    devtools::build()   # creates fidelioDiagnostics_<version>.tar.gz

#### 5) Shiny app dev loop (recommended)

    # keep a console open at the package root
    devtools::document()   # only when roxygen/exports changed
    devtools::load_all()
    fidelioDiagnostics::launch_app("diagnostic")   # or "results"

#### 6) Run the pipeline (reads config/project.yml)

    # no install needed if you only edited YAML
    fidelioDiagnostics::run_pipeline()

**Decision mini-tree**

- Changed **function code only** → `devtools::load_all()`
- Changed **roxygen/exports/docs** → `devtools::document()` then `devtools::load_all()`
- Need to **use in another/new session** → `devtools::install(upgrade = "never", dependencies = FALSE, build_vignettes = FALSE)` then restart + `library(fidelioDiagnostics)`
- About to **share** or something is flaky → `devtools::check()`
- Need an **archive** → `devtools::build()`

---

## Re-render this README

    devtools::build_readme()

> Commit both `README.Rmd` and the generated `README.md` (plus any figures under `man/figures/`).
